A pipeline to analyse time-course gene expression data

### 2019-10-08 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

The \texttt{moanin\_model} object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used or the basis or the degrees of freedom of the model.
\textbf{EAP:} You make it sound like or, but in fact the object has all of these elements, no?

@@

\textbf{EAP: Should \texttt{timely\_barplot} be a function of \texttt{moanin}?}

\textbf{NV: I'd rather not include these types of functions that are hard to test
and clutter the namespace. In order to provide reasonable visualization, they
also often need to be tweaked specfically to match the data (here, the ylim is
set specifically to the data}

\textbf{EAP: Okay, I'll leave that to you. Most of those tweaked things can just be in a \ldots{} argument and plotting functions that do annoying things are popular. I've at least rewritten it so users can more easily change those features in calls to the function rather than mucking about in the code}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{#order them by their score}
\NormalTok{cluster2Score=cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score,}\DecValTok{2}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
\NormalTok{dataToPlot=}\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{2}\NormalTok{,])}
\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

@@

First, let us run the clustering for all possible clusters of interest. We
will, for each clustering experiment, conserve (1) with within cluster sum of
squares; (2) the labels assigned to all genes. \textbf{EAP: I don't understand what you are trying to say in this last sentence. I think there's a typo}

\textbf{EAP: Also, I don't know what you mean by ``clustering experiment'' -- do you mean a clustering of the data with a particular K? Ditto for ``all possible clusters of interest'' -- do you mean all choices of K, so 5,9,11,15 in the below.}

@@

\textbf{EAP: I've increased this to be k=2-20, because you can't actually interpret these plots without a dense number of k; also you do 2-20 for stability}

@@

@@

We will run \textbf{EAP: subsampling???} method on the top 1000 genes \textbf{EAP: what is top??}, and do it \(B=20\) times for each of \(k=2-20\). We show the code below, but because of the time it takes we have evaluate these values separately and provided the results for users to explore more quickly.

@@

\textbf{EAP: It's very confusing that the headings here are C1, \ldots{} C20. It makes it seem like cluster 1, \ldots{} cluster 20, when I think (?) it is really 20 resampling results. Would be good to change that to, e.g.~B1,\ldots{},B20 or something.}

For each gene, we have an assignment to a cluster over the 20 resampling runs. For example, for \(k=5\):

\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
\hline
  & C1 & C2 & C3 & C4 & C5 & C6 & C7 & C8 & C9 & C10 & C11 & C12 & C13 & C14 & C15 & C16 & C17 & C18 & C19 & C20\\
\hline
AA204140 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 1 & 3 & 4 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
\hline
AB008911 & 5 & 5 & 1 & 4 & 1 & 2 & 3 & 3 & 2 & 3 & 3 & 3 & 4 & 4 & 5 & 5 & 2 & 4 & 1 & 4\\
\hline
AB010313 & 3 & 3 & 1 & 4 & 1 & 2 & 4 & 3 & 2 & 5 & 3 & 1 & 5 & 4 & 2 & 5 & 3 & 3 & 4 & 2\\
\hline
AB010342 & 5 & 3 & 1 & 5 & 1 & 1 & 3 & 3 & 1 & 3 & 3 & 3 & 4 & 3 & 5 & 4 & 2 & 3 & 4 & 4\\
\hline
AB011473 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
\hline
AB099518 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 4 & 4 & 1 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
\hline
\end{tabular}

@@

The model explorer strategy works as follows. For a single choice of \(k\), first perform \(n\) bootstrap
experiments to estimate the cluster centroids, followed by a step of assigning
a label to all data points. Then, choose a similarity
measure between two partitions or clusters \(S(C_1, C_2)\). \textbf{EAP: This is confusing as to whether this is a measure of similarity across different bootstrap resamples (i.e.~different sets of clusters) or across two different clusters in the same resample. Perhaps also because of confusion of headings of different resamples between labeled C1, C2, but those aren't cluster 1/2 (see comment above)} Examples are the
normalized mutual information or Fowlkes-Mallows. Finally, compute the pairwise similarity measure
between all bootstrapped partitions (i.e.~B choose 2 pairs). Repeat this procedure for different \(k\) and plot per \(k\) the cumulative density of the
obtained scores.

@@

The function \texttt{plot\_model\_explorer}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"C6"}\NormalTok{, }\StringTok{"C7"}\NormalTok{, }\StringTok{"C8"}\NormalTok{, }\StringTok{"C9"}\NormalTok{, }\StringTok{"C10"}\NormalTok{)}
\NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
\KeywordTok{plot_model_explorer}\NormalTok{(selected_labels)}
\end{Highlighting}
\end{Shaded}

@@

@@

@@

\textbf{EAP: We should again provide the code that goes with these images}

>>>>

The \texttt{moanin\_model} object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used, the basis, and the degrees of freedom of the model.

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# order them by their score}
\NormalTok{cluster2Score =}\StringTok{ }\NormalTok{cluster2Score[}\KeywordTok{order}\NormalTok{(scores_and_labels}\OperatorTok{$}\NormalTok{scores[cluster2Score, }\DecValTok{8}\NormalTok{], }\DataTypeTok{decreasing =} \OtherTok{TRUE}\NormalTok{)]}
\NormalTok{dataToPlot =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(data[cluster2Score[}\DecValTok{1}\OperatorTok{:}\DecValTok{4}\NormalTok{], ],}\StringTok{"Centroid"}\NormalTok{=kmeans_clusters}\OperatorTok{$}\NormalTok{centroids[}\DecValTok{8}\NormalTok{, ])}
\KeywordTok{plot_splines_data}\NormalTok{(dataToPlot, moanin_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{simpleY =}\OtherTok{FALSE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\textbf{Nelle: in my experience, the large the cluster is, the noisier the centroid
estimation is. I've switched to cluster 8, as this is the cluster we focus our
interpretation, and the top scoring genes seem to be to be closely related to
the centroid estimates}

@@

First, let us run the clustering for all possible clusters of interest. We
will, for each clustering, conserve (1) with within cluster sum of
squares; (2) the clustering assignement (or label) for each gene.

@@

@@

\textbf{Nelle: that was a mistake in the code. I don't actually run subsampling
experiments: only the bootstrap, and it's done in a separate file as it is too
time consuming to run in an Rmd file}

@@

We run bootstrap method on all genes differentially expressed and with a
log-fold-change higher than 2 (computed with the \texttt{lfc\_max} method), and do it
\(B=20\) times for each of \(k=2-20\). We show the code below, but because of the
time it takes we have evaluate these values separately and provided the
results for users to explore more quickly.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pvalues =}\StringTok{ }\NormalTok{de_analysis[, pval_col_to_keep]}

\NormalTok{splines_model =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{create_moanin_model}\NormalTok{(meta, }\DataTypeTok{degrees_of_freedom=}\DecValTok{6}\NormalTok{)}
\NormalTok{fishers_pval =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{pvalues_fisher_method}\NormalTok{(pvalues)}
\NormalTok{fishers_qval =}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\KeywordTok{p.adjust}\NormalTok{(fishers_pval)}

\CommentTok{# Keep only the genes found differentially expressed after combining the}
\CommentTok{# p-values using Fisher's method and with a log fold change greater than 2}
\NormalTok{genes_to_keep =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(}
\NormalTok{    lfc_max[(}\KeywordTok{rowSums}\NormalTok{(lfc_max }\OperatorTok{>}\StringTok{ }\DecValTok{2}\NormalTok{) }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{) }\OperatorTok{&}\StringTok{ }\NormalTok{(fishers_qval }\OperatorTok{<}\StringTok{ }\FloatTok{0.05}\NormalTok{), ])}
\NormalTok{y =}\StringTok{ }\KeywordTok{as.matrix}\NormalTok{(data[genes_to_keep, ])}

\CommentTok{# You may want to set the random seed of the experiment so that the results}
\CommentTok{# don't vary if you rerun the experiment.}
\CommentTok{# set.seed(random_seed)}
\NormalTok{n_genes =}\StringTok{ }\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{] }\OperatorTok{*}\StringTok{ }\NormalTok{sample_proportion}
\NormalTok{indices =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\KeywordTok{dim}\NormalTok{(y)[}\DecValTok{1}\NormalTok{], n_genes, }\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{)}

\NormalTok{random_subsample_y =}\StringTok{ }\NormalTok{y[indices, ]}
\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{::}\KeywordTok{splines_kmeans}\NormalTok{(}
\NormalTok{    random_subsample_y, splines_model, }\DataTypeTok{n_clusters=}\NormalTok{n_clusters,}
    \DataTypeTok{random_seed=}\NormalTok{random_seed,}
    \DataTypeTok{n_init=}\DecValTok{20}\NormalTok{)}

\CommentTok{# Perform prediction on the whole set of data.}
\NormalTok{kmeans_clusters =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{splines_kmeans_prediction}\NormalTok{(}
\NormalTok{    y, kmeans_clusters)}
\end{Highlighting}
\end{Shaded}

@@

Each column correspond to a bootstrap sample, each row to a gene, and each
entry to the label found for that particular clustering. Thus, for each gene,
we have an assignment to a cluster over the 20 resampling runs. For example,
for \(k=5\):

\begin{tabular}{l|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r}
\hline
  & B1 & B2 & B3 & B4 & B5 & B6 & B7 & B8 & B9 & B10 & B11 & B12 & B13 & B14 & B15 & B16 & B17 & B18 & B19 & B20\\
\hline
A\_51\_P114236 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
\hline
A\_51\_P172409 & 1 & 5 & 5 & 5 & 5 & 5 & 1 & 3 & 2 & 1 & 3 & 4 & 3 & 3 & 5 & 4 & 2 & 4 & 1 & 4\\
\hline
A\_51\_P188401 & 1 & 5 & 5 & 4 & 5 & 5 & 5 & 3 & 2 & 4 & 4 & 4 & 2 & 3 & 5 & 4 & 2 & 4 & 3 & 4\\
\hline
A\_51\_P191274 & 4 & 2 & 1 & 2 & 1 & 3 & 3 & 2 & 5 & 5 & 2 & 1 & 4 & 1 & 1 & 2 & 5 & 1 & 2 & 1\\
\hline
A\_51\_P226791 & 1 & 4 & 2 & 4 & 3 & 4 & 2 & 1 & 3 & 3 & 4 & 2 & 5 & 2 & 3 & 5 & 2 & 2 & 5 & 3\\
\hline
A\_51\_P277048 & 1 & 1 & 4 & 5 & 5 & 2 & 5 & 5 & 2 & 4 & 3 & 4 & 2 & 5 & 4 & 3 & 4 & 4 & 3 & 5\\
\hline
\end{tabular}

@@

The model explorer strategy works as follows. For a single choice of \(k\),
first perform \(n\) bootstrap experiments to estimate the cluster centroids,
followed by a step of assigning a label to all data points. Then, choose a
similarity measure between two partitions or clusters \(S(B_1, B_2)\). Examples
are the normalized mutual information or Fowlkes-Mallows. Finally, compute the
pairwise similarity measure between all bootstrapped partitions (i.e.~B choose
2 pairs). Repeat this procedure for different \(k\) and plot per \(k\) the
cumulative density of the obtained scores.

@@

The function \texttt{plot\_model\_explorer} takes in input a list of the bootstrapped
results for all labels.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{clusters =}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"B6"}\NormalTok{, }\StringTok{"B7"}\NormalTok{, }\StringTok{"B8"}\NormalTok{, }\StringTok{"B9"}\NormalTok{, }\StringTok{"B10"}\NormalTok{)}
\NormalTok{selected_labels =}\StringTok{ }\NormalTok{all_labels[clusters]}
\KeywordTok{plot_model_explorer}\NormalTok{(selected_labels)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{auc_scores =}\StringTok{ }\NormalTok{moanin}\OperatorTok{:::}\KeywordTok{get_auc_similarity_scores}\NormalTok{(all_labels)}
\KeywordTok{plot}\NormalTok{(n_clusters, auc_scores, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{type=}\StringTok{"b"}\NormalTok{, }\DataTypeTok{pch=}\DecValTok{16}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{delta_auc =}\StringTok{ }\KeywordTok{diff}\NormalTok{(auc_scores)}\OperatorTok{/}\NormalTok{auc_scores[}\DecValTok{1}\OperatorTok{:}\NormalTok{(}\KeywordTok{length}\NormalTok{(auc_scores)}\OperatorTok{-}\DecValTok{1}\NormalTok{)]}
\KeywordTok{plot}\NormalTok{(}\DecValTok{3}\OperatorTok{:}\DecValTok{20}\NormalTok{, delta_auc, }\DataTypeTok{col=}\StringTok{"black"}\NormalTok{, }\DataTypeTok{type=}\StringTok{"b"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@
